---
title: "R project"
output:
  html_document: default
  pdf_document: default
---
## Healthcare Dataset Stroke Data Analysis and Visualizations
Akshay Kapare, Dharmit Dalvi - Term project

## Dataset details

The dataset is taken from Kaggle datasets, the link for which is as follows:
https://www.kaggle.com/asaumya/healthcare-dataset-stroke-data#train_2v.csv

The dataset contains approximately 44,000 rows, and 12 columns.
Each row contains data for a single patient. The columns include attributes for patients such as their age, gender, BMI, smoking status, etc, with which we can predict the final attribute: “stroke”, which predicts if the patient might suffer from a stroke or no.

## Objective
The objective of this project is to gain insights on patients' health through analysis and visualization.

## Preprocessing
We read the dataset, our health care stroke data set and replace all the mising values in one of our columns- smoking status.
```{r}
data <- as.data.frame(read.csv("stroke_data.csv", header=T, sep=",", na.strings = c("", "NA")))
attach(data)
data$smoking_status <- as.character(data$smoking_status)
data$smoking_status <- ifelse(is.na(data$smoking_status), 
                             'No information available', data$smoking_status)

```

The frequency for genders and a barplot for the same is calculated as follows:
We have further calculated the same for only those patients who have stroke. (stroke = 1)
```{r}
#categorical 1
table(data$gender)
barplot(table(data$gender), col = "cyan", ylim = c(0,25000), las = 2, xlab = "Gender", ylab = "Stroke")
yes_stroke <- data[data$stroke == 1, ]
table(yes_stroke$gender)
barplot(table(yes_stroke$gender), col = "cyan", ylim = c(0,500), las = 2, xlab = "Gender", ylab = "Stroke")
```

We can infer that the data has maximum number of females, followed by males and others.

We have further analysed the smoking status attribute, which is a categorical variable and visualized using a pie chart:
```{r}
#categorical 2
smoking_status <- table(data$smoking_status)
slice.labels <- names(smoking_status)
slice.percents <- round(smoking_status/sum(smoking_status)*100)
slice.labels <- paste(slice.labels, slice.percents)
slice.labels <- paste(slice.labels, "%", sep = "")
pie(smoking_status, labels = slice.labels, col = hcl(c(0, 60, 120))) 
```

Percentage of people who have never smoked is the highest (37 %).

We further plot a boxplot for the age attribute and perform a five number summary:
```{r}
fivenum(data$age)
boxplot(data$age, horizontal = TRUE,  xaxt = "n", xlab = "Age of patients", col=hcl(1))
axis(side = 1, at=fivenum(data$age), labels = TRUE)
f <- fivenum(data$age)
f
```

Next, we plot a ggplot for age, with respect to stroke values (1 or 0):
We can see that for age = 80, the count for patients having stroke is maximum.
```{r}
library(ggplot2)
ggplot(data, aes(x=age)) + 
  geom_histogram(color="black", fill="lightblue") +   facet_grid(~stroke)
```

We further plot a boxplot of average glucose levels in patients that have a stroke and average glucose levels in patients that don't have stroke.
```{r}
no_stroke <- data[data$stroke == 0, ]
A <- no_stroke$avg_glucose_level
B <- yes_stroke$avg_glucose_level
boxplot(A,B, xaxt = "n", xlab = "Whether patients have stroke", ylab = "Glucose levels of patients", col=c("red", "green"))

```
The patients that don't have stroke have lower average of glucose levels, the ones that have stroke have a greater average. The range of glucose values is between ~80 to 200 for patients having stroke, whereas there are a lot of outliers in glucose levels' data for patients not having stroke.

Following is the scatterplot between three variables: age, hypertension and stroke.
```{r}
#2
dt=data.frame(Age=data$age,hypertension=data$hypertension,Stroke=data$stroke)
head(dt,10)
plot(dt , pch=20 , cex=1.5 , col="red")

```
Major inferences that we can make from the above plot are ages for patients with stroke start from 35, whereas for hypertension, majority of patients having hypertension start from the age of 15 itself.

We have further plotted a ggplot for age vs heart disease, and we have the points colored based on work_type.
```{r}
#3
ggplot(data = data) + 
  geom_point(mapping = aes(x = age, y = heart_disease, colour = data$work_type ))

```
We can infer that most patients with heart disease work in private sector.

Next we have boxplots of age, bmi and average glucose levels combined in a single plotly plot:
```{r}
library(plotly)
p <- plot_ly(data, x = data$age, type="box", name = 'age')


q <-add_trace(p, x = data$bmi, type="box", name = 'bmi')
 

w <-add_trace(q , x = data$avg_glucose_level,type = "box" , name ="glucose level" )
w
```


The pmf plot and cdf plots for age are as follows:

```{r}
#age
values <- data$age
tab <- table(values)

dframe <- as.data.frame(tab)
#dframe

x <- as.numeric(as.character(dframe$values))

# probability distribution is 
f <- dframe$Freq / (sum(dframe$Freq))

# calculate  the  mean
mu  <- sum(x * f)
mu

# variance of the  distribution is 
sigmaSquare <-  sum((x -  mu)^2 * f)
sigmaSquare

sigma <- sqrt(sigmaSquare)
sigma

plot(x, f, type = 'h',  xlab = "Age of patients", ylab = "PMF", ylim = c(0, 0.04), main = "Spike plot for Age")
abline(h = 0 )

cdf <- c(0, cumsum(f))
cdfplot <- stepfun(x, cdf)
plot(cdfplot,  verticals=FALSE, pch=16, main="CDF Plot for Age", xlab = "Age", ylab = "CDF")

```

Similar analysis of distribution for bmi is as follows:
```{r}
#bmi

values <- data$bmi
tab <- table(values)

dframe <- as.data.frame(tab)
#dframe

x <- as.numeric(as.character(dframe$values))

# probability distribution is 
f <- dframe$Freq / (sum(dframe$Freq))

# calculate  the  mean
mu  <- sum(x * f)
mu

# variance of the  distribution is 
sigmaSquare <-  sum((x -  mu)^2 * f)
sigmaSquare

sigma <- sqrt(sigmaSquare)
sigma

plot(x, f, type = 'h',  xlab = "BMI of patients", ylab = "PMF",ylim = c(0, 0.01), main = "Spike plot for BMI")
abline(h = 0 )

cdf <- c(0, cumsum(f))
cdfplot <- stepfun(x, cdf)
plot(cdfplot,  verticals=FALSE, pch=16, main="CDF Plot for BMI", xlab = "BMI", ylab = "CDF")
```

We have applied Cental Limit Theorem on age attibute as follows:
```{r}
age <-data$age

ctable <- table(age)
#ctable
mu <- mean(age)
mu
sigma <- sd(age)
sigma


dframe <- as.data.frame(ctable)
#dframe

x <- as.numeric(as.character(data$age))
#x
hist(x, probability = TRUE, xlim = c(0, 100), xlab = "age", ylab = "Density", main = "Histogram of age")

# sample size 5
samples <- 1000
sample_size <- 5
xbar <- numeric(samples)
for(i in 1:samples){
  xbar[i] = mean(sample(x, size = sample_size, replace = T))
}
hist(xbar, prob = T, xlab = "Age", 
     main = "Densities of age with sample size 5", col = "blue")
mean1 <- mean(xbar)
sd1 <- sd(xbar)
mean1
sd1
#sample size 20
samples <- 1000
sample_size <- 20
xbar <- numeric(samples)
for(i in 1:samples){
  xbar[i] = mean(sample(x, size = sample_size, replace = T))
}
hist(xbar, prob = T, xlab = "Age", 
     main = "Densities of age with sample size 20", col = "red")
mean2 <- mean(xbar)
sd2 <- sd(xbar)
mean2
sd2

#sample size 50
samples <- 1000
sample_size <- 50
xbar <- numeric(samples)
for(i in 1:samples){
  xbar[i] = mean(sample(x, size = sample_size, replace = T))
}
hist(xbar, prob = T, xlab = "Age", 
     main = "Densities of age with sample size 50", col = "green")
mean3 <- mean(xbar)
sd3 <- sd(xbar)
mean3
sd3

cat("1st distribution:\nMean =",mean1,"\nSD =",sd1)
cat("2nd distribution:\nMean =",mean2,"\nSD =",sd2)
cat("3rd distribution:\nMean =",mean3,"\nSD =",sd3)

```

Similarly, We have applied Cental Limit Theorem on average glucose level attibute as follows:

```{r}
glucose <-data$avg_glucose_level  

ctable <- table(glucose)
#ctable
mu <- mean(glucose)
mu
sigma <- sd(glucose)
sigma


dframe <- as.data.frame(ctable)
#dframe

x <- as.numeric(as.character(data$avg_glucose_level))
#x
max(data$avg_glucose_level)

hist(x, probability = TRUE, xlim = c(0,300 ), xlab = "Average Glucose Level", ylab = "Density", main = "Histogram of Average Glucose Level")
#sample size 5
samples <- 1000
sample_size <- 5
xbar <- numeric(samples)
for(i in 1:samples){
  xbar[i] = mean(sample(x, size = sample_size, replace = T))
}
hist(xbar, prob = T,xlim = c(0,200), xlab = "Avg Glucose level", 
     main = "Densities of Average Glucose Level with sample size 5", col = "blue")
mean1 <- mean(xbar)
sd1 <- sd(xbar)
mean1
sd1
#sample size 20
samples <- 1000
sample_size <- 20
xbar <- numeric(samples)
for(i in 1:samples){
  xbar[i] = mean(sample(x, size = sample_size, replace = T))
}
hist(xbar, prob = T,xlim = c(0,200), xlab = "Average Glucose Level", 
     main = "Densities of Average Glucose Level with sample size 20", col = "red")
mean2 <- mean(xbar)
sd2 <- sd(xbar)
mean2
sd2

#sample size 50
samples <- 1000
sample_size <- 50
xbar <- numeric(samples)
for(i in 1:samples){
  xbar[i] = mean(sample(x, size = sample_size, replace = T))
}
hist(xbar, prob = T,xlim = c(0,200), xlab = "Average Glucose Level", 
     main = "Densities of Average Glucose Level with sample size 20", col = "green")
mean3 <- mean(xbar)
sd3 <- sd(xbar)
mean3
sd3

cat("1st distribution:\nMean =",mean1,"\nSD =",sd1)
cat("2nd distribution:\nMean =",mean2,"\nSD =",sd2)
cat("3rd distribution:\nMean =",mean3,"\nSD =",sd3)

```
In both cases, our mean remains almost the same, but as sample size increases, standard deviation decreases, proving the applicability of Central Limit Theorem on both of our attributes.

We further performed sampling on age attributes of our data:

```{r}
#sampling 
library(sampling)
table(data$age)
hist(data$age)
mean_without_sampling <- mean(data$age)
sd(data$age)

```

Following are various sampling methods for sample size 500:

```{r}
#sample size = 500
# srswor
sample.size <- 500
s <- srswor(sample.size,nrow(data))
sample.1 <- data[s != 0, ]

mean_srswor <- mean(sample.1$age)
#srswr
set.seed(153)
s <- srswr(sample.size, nrow(data))
sample.2 <- data[s != 0, ]
mean_srswr <- mean(sample.2$age)

```

```{r}
#Systematic Sampling
N <- nrow(data)
n <- 1000

k <- ceiling(N / n)
k

r <- sample(k, 1)
r

s <- seq(r, by = k, length = n)
#s

sample.3 <- data[s, ]
table(sample.3$age)
mean_systematic <- mean(sample.3$age)


```

```{r}
#Systematic Sampling with unequal probabilities
pik <- inclusionprobabilities(data$age, sample.size)
s <- UPsystematic(pik)
sample.4 <- data[s != 0, ]
table(sample.4$age)

data["age_range"] = NA

data$age_range <- cut(data$age, breaks = c(0, 25, 50, 75, Inf), labels = c('A', 'B', 'C', 'D'))

data_age <- data.frame(
  age = data$age,
  age_range = data$age_range
)

freq <- table(data_age$age_range)
freq

set.seed(123)
head(data_age)
st.sizes <- sample.size * freq / sum(freq)

st.1 <- strata(data_age, stratanames = c("age_range"),
               size = st.sizes, method = c("srswor"),
               description = TRUE)

#st.1

st.sample1 <- getdata(data, st.1)

#st.sample1

```

```{r}
# cluster sampling
cl <- cluster(data, c("age"), size = 4, method = "srswor")
sample.6 <- getdata(data, cl)

table(sample.6$age)

mean_cluster <- mean(sample.6$age)

```

For sample size 500, the means through all sampling methods' samples are as follows:
```{r}
#sample size 500
mean_without_sampling
mean_srswor
mean_srswr
mean_systematic
mean_cluster
```

Following are various sampling methods for sample size 1000:

```{r}
#sample size = 1000
# srswor
sample.size <- 1000
s <- srswor(sample.size,nrow(data))
sample.1 <- data[s != 0, ]

mean_srswor <- mean(sample.1$age)
#srswr
set.seed(153)
s <- srswr(sample.size, nrow(data))
sample.2 <- data[s != 0, ]
mean_srswr <- mean(sample.2$age)

```

```{r}

#Systematic Sampling
N <- nrow(data)
n <- 1000

k <- ceiling(N / n)
k

r <- sample(k, 1)
r

s <- seq(r, by = k, length = n)
s

sample.3 <- data[s, ]
table(sample.3$age)
mean_systematic <- mean(sample.3$age)

```

```{r}
#Systematic Sampling with unequal probabilities
pik <- inclusionprobabilities(data$age, sample.size)
s <- UPsystematic(pik)
sample.4 <- data[s != 0, ]
table(sample.4$age)

```

```{r}
# Stratified sampling
freq <- table(data_age$age_range)
freq

set.seed(123)
head(data_age)
st.sizes <- sample.size * freq / sum(freq)

st.1 <- strata(data_age, stratanames = c("age_range"),
               size = st.sizes, method = c("srswor"),
               description = TRUE)

#st.1

st.sample1 <- getdata(data, st.1)

#st.sample1

```

```{r}
# cluster sampling
cl <- cluster(data, c("age"), size = 4, method = "srswor")
sample.6 <- getdata(data, cl)

table(sample.6$age)

mean_cluster <- mean(sample.6$age)
```

Following are the mean values for all of our sampling methods:
```{r}

#sample size 1000
mean_without_sampling
mean_srswor
mean_srswr
mean_systematic
mean_cluster
```

Following is an R code for finding if a patient with an input ID has a stroke or no. (Please refer to the .R file for code to dynamically insert any id as input)
```{r}
id <- 35327
id_stroke <- data[data$id == id, ]
if(id_stroke$stroke == 0){
  print(paste("The patient with ID", id, "does not have a stroke"))
} else{
  print(paste("The patient with ID", id, "has a stroke"))
}
```

Following is the word cloud for work type:
```{r}
# Word cloud for work type
# Install
#install.packages("tm")  # for text mining
#install.packages("wordcloud") # word-cloud generator 
#install.packages("RColorBrewer") # color palettes
# Load
library("tm")
library("SnowballC")
library("wordcloud")
library("RColorBrewer")
yes_stroke <- data[data$stroke == 1, ]

table(yes_stroke$work_type)

workCorpus <- Corpus(VectorSource(yes_stroke$work_type))
workCorpus <- tm_map(workCorpus, content_transformer(tolower))
wordcloud(workCorpus, max.words = 100, random.order = FALSE)
```